\section{Related Work}
The work on decidable fragments of logics with recursively defined functions that can be used to verify heap-manipulating programs runs quickly into the problem of finding decidable fragments of separation logic \cite{Reynolds2002} with inductive predicates, upon which a lot of work has been done. While the first works such as  \cite{BerdineCalcagnoO'Hearn2004, BerdineCalcagnoO'Hearn2005, CookHaaseChristoph2011,PerezAntonioRybalchenko2011 } handled only list segments, there has been further improvement. \cite{PerezRybalchenko2013} hanldes only list segments, but allows conjunctions with pure formulae from arbitrary decidable SMT theories. \cite{BrotherstonFuhsPerez2014} argues decidablility for separation logic with arbitrary/user-defined inductive predicates, and \cite{LeTatsutaSun2017,LeSunChin2016 } adapts the idea of using equisatisfiable `base' formulas instead of inductive predicates to include arithmetic constraints satisfying a certain property. However, none of these would be directly applicable to program verification since they do not address entailment, and do not contain an implication in their logic.  \cite{PerezRybalchenko2013} does contain implications, but it does not allow for spatial formulas other than points-to, list segment and in particular not provide support for multiple natural measures which our work provides. However, in the light of our broader work on context-logics and delta-logics, it is valuable to consider the satisfiability problem in the context-logic since it models the unchanging and does not need to address entailments, of which these works above provide useful decision procedures.\\
In constrast, \cite{PiskacWiesZufferey2014, PiskacWiesZufferey2014Tool} do address program verification and provides a decidable logic on spatial formulas that when reduced to SMT can be extended with other decidable SMT theories. There is reasonable overlap in the spirit of this claim to our work, where allowing expressions in arbitrary decidable SMT theories on $\Delta{}$  would preserve decidability as well, and would fit well within the paradigm of delta-logics. But where \cite{PiskacWiesZufferey2014} can support trees/tree segments, it does not provide support for measures. \cite{MadhusudanParlatoQiu2011, IosifRogalewiczSimacek2013} differ similarly \textbf{maybe add more? Where to I place \cite{BouajjaniDruagoiEnea2009} ?}.\\
Lastly there is work in the spirit of \cite{ChinDavidNguyen2007, QiuGargStefanescu2013}, where one dispenses with decidability and are therefore incomparable in that respect with our work. In particular, \cite{QiuGargStefanescu2013} is a powerful method capable of verifying all of our examples, but such logics are incomplete and cannot give meaningful counterexamples like our work does.